How environmental drivers of spatial synchrony interact

### 2021-07-30 Paper/Paper.Rnw
bit more progress on the theory section of main text

%Alternative titles
%Interactions among Moran effects
%Interacting Moran effects
%Interactions between environmental drivers of spatial synchrony

@@

%The following is where you load in the numeric results that will be embedded in the text
<<eval=T,echo=F,message=F,warning=F>>=
#place R code here for loading in necessary variables from Results
@

@@

%Tentative paragraph outline written on my phone on 3/20/2021
%1) Importance of synchrony
%2) Classic problem of identifying causes of synchrony, but recently largely resolved due to new methods
%  -elaborate briefly on a couple methods
%3) in fact, it has recently been discovered that environmental causes of synchrony can interact, so that ...
%  -related to, but distinct from, interaction between environmental and dispersal causes of synchrony discussed, for instance by Kendal et al
%  -make sure you describe the two cases, cite the earlier papers, mention a few words about the two systems
%4) the simplest form of interacting moran effects can be demonstrated when a simple model, though elaboration of the true nature of this mechanism %will take the rest of this manuscript
%  -use Moran model
%  -but with two drivers
%  -moran theorem says correlation of pops equals correlation of sum of drivers, so work it out
%5) general idea of two noises lagged or time delayed relative to each other and their effects oppositely lagged - this may be where a pedagogical %fig comes in, or maybe it is just worsd here and then the pedagog fig comes in later, at end of theory section, in form of examples.
%6) paragraph on why we would expect interacting moran effects to be common, return to this point in the discussion
%7) final para on goals, manuscript structure, etc.
%  -G1) Elaborate a mathematical/statistical theory of interacting Moran effects which both allows both a detailed understanding 
%      of how the mechanism works and fecilitates statistical investigation of whether the effect is present in empirical systems.
%  -G2) Apply the theory and associated stats to revisiting kelp, the intention being to further elaborate the mechanism itself
%  -G3) Apply the theory and associated stats to plankton, the intention being to further elaborate the mechanism itself
%
%Notes:
%1) You need to make it clear the earlier papers already identified interacting Moran effects, so what are we doing?
%We are exploring theoreitically what the mechanism is, here, and how general it might be, and what different forms it 
%might take, using the theory, and also using the theory to further illuminate the two examples. The emphasis here
%is on illuminating the general mechanism of interacting Moran effects through the theory and its application, whereas
%the earlier papers focussed on the specific cases. The Fourier theory here serves to illuminate the general mechanism,
%whereas the earlier papers used wavelet statistics applied directly to data and did not really explore the general 
%mechanism. Need to work some of these thoughts in without seeming defensive.

@@

%Tentative broad outline:
%DONE 1) Intro spectral quantities, the spectrum, the cospectrum, and the cross spectrum of two time series/processes. And the spectral
%matrix for putting them all in one structure.
%2) Define your measures of timescale/frequency-specific synchrony and crosss-variable synchrony, explain why these make sense as
%ways to measure those things.
%3) Present the model and the theory
%4) Present "toy" examples and fig 1 (need to think a lot about these)
%
%Notes:
%Compared to the deer paper, this is a much deeper presentation of the theory in the main text. I'm not sure that's avoidable.
%For one thing, the theory is explicitly applied to get the results, later, and that was not the case for the deer paper.

\noindent Our main theoretical work requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series.
If $w_i(t)$ is a stochastic process or time series measured at location $i$, the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over the timescale, or period, of the oscillation. 
The spectrum, $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates more. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics at $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=1/5 \text{y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur on different timescales, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are strong, and are strongly correlated, i.e., they have the same or similar phase. 
(Here, $j$ refers to another location where measurements were performed.) The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i$ and $w_j$ are strong; and the complex phase of the cross spectrum at $f$ then quantifies
the phase lag between those oscillations. The notation $S_{ww}$ refers to the \emph{spectral matrix}, a 
matrix for which the $ij^{\text{th}}$ entry is $S_{w_i w_j}$. 
Spectral methods are standard tools in many fields of science, 
including ecology \citep{Vasseur2007,Reuman2008,Defriez2017}, and many background references are available 
[e.g., \cite{Brillinger2001}].

>>>>

%Alternative titles
%Interactions among Moran effects
%Interacting Moran effects
%Interactions between environmental drivers of spatial synchrony
%Environmental drivers of spatia synchony can strongly interact

@@

%The following is where you load in the numeric results that will be embedded in the text
%<<eval=T,echo=F,message=F,warning=F>>=
%#place R code here for loading in necessary variables from Results
%@

@@

%Tentative paragraph outline written on my phone on 3/20/2021
%1) Importance of synchrony
%2) Classic problem of identifying causes of synchrony, but recently largely resolved due to new methods
%  -elaborate briefly on a couple methods, mention the importance of a timescale-resolved appproach
%3) in fact, it has recently been discovered that environmental causes of synchrony can interact, so that ...
%  -related to, but distinct from, interaction between environmental and dispersal causes of synchrony discussed, for instance by Kendal et al
%  -make sure you describe the two cases, cite the earlier papers, mention a few words about the two systems
%4) the simplest form of interacting moran effects can be demonstrated when a simple model, though elaboration of the true nature of this mechanism %will take the rest of this manuscript
%  -use Moran model
%  -but with two drivers
%  -moran theorem says correlation of pops equals correlation of sum of drivers, so work it out
%5) general idea of two noises lagged or time delayed relative to each other and their effects oppositely lagged - this may be where a pedagogical %fig comes in, or maybe it is just worsd here and then the pedagog fig comes in later, at end of theory section, in form of examples.
%6) paragraph on why we would expect interacting moran effects to be common, return to this point in the discussion
%7) final para on goals, manuscript structure, etc.
%  -G1) Elaborate a mathematical/statistical theory of interacting Moran effects which both allows both a detailed understanding 
%      of how the mechanism works and fecilitates statistical investigation of whether the effect is present in empirical systems.
%  -G2) Apply the theory and associated stats to revisiting kelp, the intention being to further elaborate the mechanism itself
%  -G3) Apply the theory and associated stats to plankton, the intention being to further elaborate the mechanism itself
%
%Notes:
%1) You need to make it clear the earlier papers already identified interacting Moran effects, so what are we doing?
%We are exploring theoreitically what the mechanism is, here, and how general it might be, and what different forms it 
%might take, using the theory, and also using the theory to further illuminate the two examples. The emphasis here
%is on illuminating the general mechanism of interacting Moran effects through the theory and its application, whereas
%the earlier papers focussed on the specific cases. The Fourier theory here serves to illuminate the general mechanism,
%whereas the earlier papers used wavelet statistics applied directly to data and did not really explore the general 
%mechanism. Need to work some of these thoughts in without seeming defensive.

@@

%Tentative broad outline:
%DONE 1) Intro spectral quantities, the spectrum, the cospectrum, and the cross spectrum of two time series/processes. And the spectral
%matrix for putting them all in one structure.
%DONE 2) Define your measures of timescale/frequency-specific synchrony and crosss-variable synchrony, explain why these make sense as
%ways to measure those things.
%3) Present the model and the theory
%4) Present "toy" examples and fig 1 (need to think a lot about these)
%
%Notes:
%Compared to the deer paper, this is a much deeper presentation of the theory in the main text. I'm not sure that's avoidable.
%For one thing, the theory is explicitly applied to get the results, later, and that was not the case for the deer paper.

\noindent Our main theoretical work requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series.
If $w_i(t)$ is a stochastic process or time series measured at location $i$ (e.g., the population density time series
of a species of interest in that location), the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over the timescale, or period, of the oscillation. 
The spectrum $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates more. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics of $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=1/5 \text{ y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur on different frequencies, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are both strong and strongly correlated, i.e., they have the same or similar phase. 
(Here, $j$ refers to another location where measurements were performed.) The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i(t)$ and $w_j(t)$ are both strong; and the complex phase of the cross spectrum at $f$ then quantifies
the phase lag between those components. The notation $S_{ww}(f)$ refers to the \emph{spectral matrix}, 
the $ij^{\text{th}}$ entry of which is $S_{w_i w_j}(f)$. Spectral methods are standard tools in many fields, 
including ecology [e.g., \cite{Vasseur2007,Reuman2008,Defriez2017}], and many background references are available 
[e.g., \cite{Brillinger2001}].

Much prior work demonstrates the importance of a frequency-specific approach to
synchrony [e.g., \cite{Vasseur2007,Keitt2008,Defriez2016,Sheppard2016,Desharnais2018,Anderson2021}], 
and it will turn out (see below, and \emph{Discussion}) that a frequency-specific 
approach is essential to the development of our new theory. 
%***DAN: Good topic for a Discussion para, the fact that this could not really be done without taking frequency specificity into account, and why, though need to think a bit more before writing
We therefore here define, using the spectral methods outlined above, a frequency-specific measure of spatial synchrony, 
as well as a new concept of \emph{cross-variable spatial synchrony}. If time series data $w_i(t)$ for $t=1,\ldots,T$ 
were gathered at locations $i=1,\ldots,N$, our spatial synchrony measure is simply
$\rho_{ww}=\left( \sum_{i \neq j} S_{w_i w_j} \right)/(N^2-N)$, the average of the cross spectra for all pairs
of distinct locations. This is a real-valued function of frequency, the integral (across frequencies) of which
is the classic, non-frequency-specific measure $\left( \sum_{i \neq j} \cov(w_i,w_j) \right)/(N^2-N)$ (see SI section
\ref{SM-Sect:measuring_sync} for details, here). If two time series $\epsilon_i^{(1)}(t)$ and $\epsilon_i^{(2)}(t)$
($t=1,\ldots,T$ and $i,j=1,\ldots,N$) were measured at each sampling location (e.g., two environmental variables),
cross-variable spatial synchrony (or, simply, \emph{cross synchrony}) between the variables
is $\rho_{\epsilon^{(1)}\epsilon^{(2)}}=\left( \sum_{i \neq j} S_{\epsilon_i^{(1)}\epsilon_j^{(2)}} \right)/(N^2-N)$. 
This is conceptually interpretable as pertaining to spatial synchrony because it makes comparisons across 
distinct locations, i.e., $i \neq j$. 
It is interpretable as cross-variable synchrony because comparisons are between the two variables, i.e., cross spectra 
between time series components of $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are used. The new index 
takes into account possible time lags. For instance, if both $\epsilon^{(1)}$ and $\epsilon^{(2)}$ show strong $4$-year-period 
oscillations, but peaks in the $\epsilon^{(2)}$ oscillations consistently lag peaks in the $\epsilon^{(1)}$ oscillations 
by a year, then $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$, which is complex valued, will have high magnitude at
frequency $f=1/4 \text{ y}^{-1}$, and will have  phase at that frequency equal to $\pi/2$, reflecting the one-year 
lag between the variables.
%***DAN: make sure I'm not making a sign error here when you create the additional examples for sup mat
See SI section \ref{SM-Sect:measuring_sync} for further examples and explanations of the new cross synchrony index.

### 2021-07-30 Paper/SupMat.Rnw
started writing theory section of the main text
bit more progress on the theory section of main text

%new commands and so on
\newcommand{\mcP}{\mathcal{P}}
\newcommand{\mcB}{\mathcal{B}}
\newcommand{\Rp}{\text{Re}}
\newcommand{\cov}{\text{cov}}
\let\conj\overline

@@

%***Tentative author order
\author[a,b]{Daniel C. Reuman}
\author[c]{Max Castorani}
\author[d,e]{Tom Bell}
\author[f]{Kyle Cavanaugh}
\author[c]{Jonathan Walter}
\author[g]{Lawrence W. Sheppard}

\affil[a]{Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}
\affil[c]{University of Virginia}
\affil[d]{WHOI}
\affil[e]{UCSB}
\affil[f]{UCLA}
\affil[g]{MBA}

>>>>

%new commands and so on
\newcommand{\mcP}{\mathcal{P}}
\newcommand{\mcB}{\mathcal{B}}
\newcommand{\Rp}{\text{Re}}
\newcommand{\var}{\text{var}}
\newcommand{\cov}{\text{cov}}
\let\conj\overline

@@

%***Tentative author order
\author[a,b]{Daniel C. Reuman}
\author[c]{Max Castorani}
\author[d,e]{Tom Bell}
\author[f]{Kyle Cavanaugh}
\author[c]{Jonathan Walter}
\author[a,g]{Lawrence W. Sheppard}

\affil[a]{Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}
\affil[c]{Department of Environmental Sciences, University of Virginia}
\affil[d]{Applied Ocean Physics and Engineering Department, Woods Hole Oceanographic Institution}
\affil[e]{Earth Research Institute, University of California, Santa Barbara}
\affil[f]{Department of Geography, University of California, Los Angeles}
\affil[g]{Marine Biological Association of the United Kingdom}
